Explainable AI with Probabilistic Graphical Models

### 2019-11-13 conclusions/future-work.tex
Thesis complete

\section{Future Work} \label{sec:future-work}
When considering possible future work, we need to distinguish between tasks whose scope is to \textit{address limitations of the current methods} and those related to \textit{expansion of the current work} or \textit{novel applications for it}.

\subsection{Addressing Limitations of Current Work}
\subsubsection{System}
Regarding the first class of future work, the system developed in this thesis suffers from the limitations inherent to any proof of concept software, namely a general lack of polish and of usability.
The methods themselves are studied to be able to surface explanations in the clearest way possible, but substituting the console-based frontend for a GUI, local or on a web-server, would certainly lead to a marked improvement in the user experience.
Also related to the experimental nature of the software is the fact that it was not built from the start with a coherent end-goal but was extended in a non-organic manner as more and novel methods were selected for exploration.
This has necessarily lead to the implementation being somewhat fragmented and rich in \enquote{workarounds}.
The choice of Python as the implementation language, while definitively advantageous for rapid development thanks to the comprehensive set of data science and machine learning libraries available, brought with it severe limitations.
A Python application, while portable, does not provide a \textit{native} experience on any platform; the current project also leaned heavily on the \textit{Anaconda} package manager\footnote{\url{https://www.anaconda.com}} so any user wishing to use it on their machine would need to deal with a potentially complex setup process.
A better alternative to a full rewrite in a compiled language would be a web GUI to a Python backend\footnote{For example \url{https://github.com/epeios-q37/atlas-python}} that would enable portability without requiring to completely change the implementation.
In any case, a rewrite of the application that dropped many redundancies, unused code and non-user-focused features, would be desirable.

\subsubsection{Methodology}
Based on the \enquote{formal} (see Subsection \ref{subsec:explainability-validation-results}) and \enquote{informal} (see Section \ref{sec:implemented-tool}) feedback received by the Istituto Cantonale di Patologia (ICP), it appears that the \enquote{dialogue} interaction modes (see Subsection \ref{subsec:dialogue-results}) should be modified, as the experts were not able to \enquote{instinctively} understand their workings, even after having perceived their potential.
The experts believed that with extra time they would have been able to use this interaction mode fruitfully, but this is a symptom of a failure on the part of the software as a system designed to be explainable should definitively require as little effort from the user as possible.
The ICP has nonetheless confirmed its intention to continue using the software, focusing in particular on the dialogue modes of interaction, as they feel they have potential as research tools.

The evaluation methodology borrowed methods from the Social Sciences but could certainly be improved by experts in this domain, who will undoubtedly be better versed in the methodological details compared to the author of this thesis, whose academic background is firmly in Computer Science.

\subsubsection{Additional Evaluations}
A more extended evaluation period would certainly be a good thing as it would also enable the assessment of the effect of \textit{novelty} of certain interaction modes and help in factoring out the \textit{experts' preconceptions} regarding what an explanation should look like (see Subsection \ref{subsec:domain-experts-initial-expectations}).

The \enquote{Pseudo-MPE} query also presented elements of unclearness for the experts, as they were confused on the non-monotonicity of the probability of the elements in the deduction chain (see Subsection \ref{subsec:results-pseudo-mpe-query}).
This is certainly a point to investigate further by implementing and evaluating alternative \textit{output modalities}, maybe even \textit{verbal} ones as these were seen to be preferred by the clinicians over all others.
The objective would be to identify the \textit{point of attrition} and discriminate if this were the actual method or only the way its outputs were displayed.
In the same vein, the reasons for the experts' misunderstanding in how to implement questions 23, 25, 27 on the system (see Section \ref{sec:results-validation-results}) should also be investigated more thoroughly.

\subsection{Extensions and Novel Applications}
The second class of future work is that concerning the expansion of the currently developed techniques.
DAOOPT was unfortunately not used as a benchmark for the \enquote{pseudo-MPE} algorithm because of questions regarding the reliability of the MPE solutions obtained by using it (see Subsection \ref{subsec:results-mpe-calculation-issues}).
There is, however, a roadmap to follow up in more detail on the evaluation of the \enquote{pseudo-MPE} algorithm in a separate paper co-authored with IDSIA researchers.

One of the issues encountered in this thesis was the late removal of three variables because of their lack of clinical significance, as described in Subsection \ref{subsec:removal-clinical-variables}.
The ICP is in possession of a newer data set, homogeneous to the \textit{benchmark data set} used throughout this thesis (see Section \ref{sec:data-set}), where the values of two of the three variables (\enquote{FISH} and \enquote{loss17}) are defined and thus have clinical noteworthiness.
The third variable (\enquote{mut17q21}) is still missing too high a number of values to be relevant but these could be predicted using the BN or other discriminative ML techniques, as there is a real \textit{causal} dependency between it and the other two - as \enquote{mut17q21}, the mutation on chromosome 17, is identified through the technique of fluorescent in situ hybridization (FISH).
These are all novel clinical variables and they are thus open to being the subject of many research questions; the ICP has confirmed its interest in using the tool developed in this thesis to pursue such investigations that could, potentially, lead to scientific publications.

Connected to the previous point, the ICP suggested to develop a \enquote{workflow} for the importing of new data sets, distinct from the one used throughout this thesis.
The experts also felt that functionalities to save the textual and visual outputs of previous queries, to save the state of a \enquote{dialogue} in order to resume it from a certain point onwards and ready-made query masks could undoubtedly be useful.

The ICP has also confirmed that it will be investigating the clinical relevance, through literature review and experiments, of some of the results obtained while using the tool; in particular, these would be those related to:
\begin{itemize}
  \item the lack of correlation between common clinical pathological features (i.e., morphology, TNM, grade, hormonal status) with age at diagnosis;
  \item the correlation between positive progesterone expression and low tumour proliferation index (ki67);
  \item the correlation between negative oestrogen receptor expression and high tumour proliferation index (ki67).
\end{itemize}

Finally, some interesting research avenues were not explored, for example a work by \citet{Kyrimi2016} who introduced a method that can be seen as the \enquote{inverse} of that proposed in \citep{Butz2018}; instead of looking for the outcome best explained by the given evidence, the authors propose techniques to find the evidence that best explains target variables.
Adapting this \enquote{inverse} method could potentially extend the \textit{explanatory powers} of the system, for example by enabling clinicians to understand which variables best justify a series of observed features in a patient.
It should not be too difficult as an undertaking, as the current Python implementation is highly modular, by virtue of being based on open-source data structure implementations.

>>>>

\section{Future Work} \label{sec:future-work}
When considering possible future work, one needs to distinguish between tasks whose scope is to \textit{address limitations of the current methods} and those related to \textit{expansion of the current work} and \textit{novel applications for it}.

\subsection{Addressing Limitations of Current Work}
\subsubsection{Methodology}
Based on the \enquote{formal} (see Subsection \ref{subsec:explainability-validation-results}) and \enquote{informal} (see Section \ref{sec:implemented-tool}) feedback received by the medical partner, it appears that the \enquote{dialogue} interaction modes (see Subsection \ref{subsec:dialogue-results}) should be modified, as the experts were not easily able to understand their workings, even after having perceived their potential.
The experts believed that with extra time they would have been able to use this interaction mode productively, but this is a symptom of a failure on the part of the software as a system designed to be explainable should definitively require as little effort from the user as possible.
The ICP has nonetheless confirmed its intention to continue using the software, focusing in particular on the dialogue modes of interaction, as they feel they have potential as research tools.
The evaluation methodology borrowed techniques from the social sciences but could undoubtedly be improved by experts in this domain, who will certainly be better versed in the methodological details compared to the author of this thesis, whose academic background is firmly in computer science.

\subsubsection{Additional Evaluations}
A more extended evaluation period could certainly be recommended as it would also enable the assessment of the effect of \textit{novelty} of certain interaction modes and help in factoring out the \textit{experts' preconceptions} regarding what an explanation should look like (see Subsection \ref{subsec:domain-experts-initial-expectations}).

The \enquote{pseudo-MPE} query also presented elements of ambiguity for the experts, as they were confused on the non-monotonicity of the probability of the elements in the deduction chain (see Subsection \ref{subsec:results-pseudo-mpe-query}).
This is unquestionably a point to investigate further by implementing and evaluating alternative \textit{output modalities}, as well as potentially \textit{linguistic} ones as these were proved to be preferred by the clinicians over all others.
The objective would be to identify the \textit{point of attrition} and discriminate if it were to be found in the actual underlying method or only in the way its outputs were displayed.
In the same vein, the reasons for the experts' misunderstanding in how to implement questions 23, 25, 27 on the system (see Section \ref{sec:results-validation-results}) should also be investigated more thoroughly.

\subsubsection{System}
The system developed in this thesis is bound by the limitations inherent to any proof of concept software, namely a general lack of polish and of somewhat lacking usability.
The methods themselves are studied to be able to surface explanations in the clearest way possible, but substituting the console-based frontend for a GUI - local or web-based - would beyond doubt lead to a marked improvement in the user experience.
Also related to the experimental nature of the software is the fact that it was not built from the start with a coherent end-goal but was extended in a non-organic manner as new methods were selected for exploration or novel ones were developed.
As a result the implementation presents some fragmentation and is rich in \enquote{workarounds}.
The choice of Python as the implementation language, while definitively advantageous for rapid development thanks to the comprehensive set of data science and machine learning libraries available, brought with it some clear limitations.
A Python application, while portable across different systems, does not provide a \textit{native} experience on any platform; the current project also relied heavily on the \textit{Anaconda} package manager\footnote{\url{https://www.anaconda.com}} so any user wishing to use the tool on their machine would need to deal with a potentially intricate setup process.
A better alternative to a full rewrite in a compiled language would be a web GUI to a Python backend\footnote{For example by using \url{https://github.com/epeios-q37/atlas-python}.} that would enable portability without requiring to completely change the implementation.
Nevertheless, a rewrite of the application that dropped many redundancies, unused code, and non-user-focused features, is a necessary part of future work.

\subsection{Extensions and Novel Applications}
The second class of future work concerns the expansion of the current techniques.
During this research it was not possible to utilise the software tool DAOOPT (see Subsection \ref{subsec:libraries} under the \enquote{DAOOPT} header) for exact MPE inferences and it was consequently not used as a benchmark for the \enquote{pseudo-MPE} algorithm (see Subsection \ref{subsec:results-mpe-calculation-issues}).
There is, however, a roadmap to follow up in more detail on the evaluation of the \enquote{pseudo-MPE} algorithm in a separate paper co-authored with IDSIA researchers.

One of the issues encountered in this thesis was the late removal of three variables due to their lack of clinical significance, as described in Subsection \ref{subsec:removal-clinical-variables}.
The ICP is in possession of a newer data set, homogeneous to the \textit{benchmark data set} used throughout this thesis (see Section \ref{sec:data-set}), where the values of two of the three variables (\enquote{FISH} and \enquote{loss17}) are defined and thus have clinical noteworthiness.
The third variable (\enquote{mut17q21}) is still missing too high a number of values to be relevant but these could be predicted using the BN or other discriminative ML techniques.
The estimation of the values of this last variable is bound to be accurate, as there is a real \textit{causal} dependency between it and the other two; this is because \enquote{mut17q21}, the mutation on chromosome 17, is identified through the technique of \textit{fluorescent in situ hybridisation} and is thus also tightly coupled with the results of \enquote{loss17}.
These are all novel clinical variables and they are thus open to being the subject of many research questions; the medical partner has confirmed its interest in using the tool developed in this thesis to pursue such investigations and this could, potentially, lead to scientific publications.

Connected to the previous point, the expert users suggested developing a \enquote{workflow} for the importing of new data sets, thus extending the clinical capabilities of the tool.
The experts also felt that the ability to save the textual and visual outputs of previous queries could undoubtedly be useful, together with the capacity to \enquote{snapshot} the state of a \enquote{dialogue} in order to resume it from a certain point onwards.
Ready-made query masks that could reduce the time needed to execute similar questions were also a request.

The ICP has also confirmed that it will be investigating the clinical relevance, through literature reviews and experiments, of some of the results obtained while using the tool; in particular, these would be those related to:
\begin{itemize}
  \item the lack of correlation between common clinical pathological features (i.e., morphology, TNM, grade, hormonal status) with age at diagnosis;
  \item the correlation between positive progesterone expression and low tumour proliferation index (ki67);
  \item the correlation between negative oestrogen receptor expression and high tumour proliferation index (ki67).
\end{itemize}

Finally, some interesting research avenues were not explored, for example a work by \citet{Kyrimi2016} who introduced a method that can be seen as the \enquote{inverse} of that proposed by \citet{Butz2018}.
Instead of looking for the outcome best explained by the given evidence, this other paper proposes techniques to find the evidence that best explains the chosen target variables.
Adapting this \enquote{inverse} method could potentially extend the \textit{explanatory powers} of the system, for example by enabling clinicians to understand which variables best justify a series of observed features in a patient.
It should not be too difficult an undertaking, as the current Python implementation is highly modular, by virtue of being based on standard open-source data structure libraries.

### 2019-11-13 introduction/context.tex
Thesis complete

\section{Context} \label{sec:intro-context}
While Artificial Intelligence (AI) - as a field - has existed for nearly seventy years \citep{moor2006dartmouth}, the concept of artificial intelligence dates back to at least Ancient Greece.  
In ancient times, artificial intelligence embodied in the dream of mechanical men was part of the domain of myth; in the twentieth century, of that of science.  
During this last decade, Artificial Intelligence is not anymore a part of a single specific domain, as it has materialised out of Man's imagination, broken out of laboratories and has been given lease to act in the world at large.

Artificial Intelligence and specifically Machine Learning, the branch of AI that specialises in creating computer systems able to learn their programming from real-world examples instead of having to be explicitly coded by a human, have in the last two decades seen incredible success\footnote{See \citet{shalev2014understanding} for an introduction to Machine Learning and Figure \ref{fig:ai-overview} to get an overview of the field of AI}.
No sector of our economy has been left untouched by the recent and rapid rise of machine learning that has been enabled by the rediscovery of neural networks, the availability of Big Data and increased availability of parallel computing power\footnote{A compelling talk, explaining the recent rise of ML, by AI pioneer Geoff Hinton can be found at \citet{hintonmachinelearning}}.
Fields as diverse and as critical as are government, healthcare, finance and bioinformatics have been revolutionised and the possibility has been set for new ones - such as self-driving vehicles - to be born\footnote{For an overview of the impacts of AI on current society and labour market see \citet{schwab2017fourth}}.
The ever increasing reliance of our society on ever more complex machine learning-driven algorithms can only make us worry ever more about the ethical problems posed by such a situation.  
Our society has only very recently been confronted with the dilemma of assigning blame when a driverless car causes the death of a person\footnote{See \citet{uberkillspedestrian}}, but this moral issue is only the tip of the iceberg, even when focusing only on the automotive industry.  
As more and more decisions are made in an automated way, with many of them significantly impacting both individuals and society at large, it comes natural to stop and wonder what are the characteristics we would want the systems tasked with making these decisions to possess.

 \begin{figure}[htbp]
\centerline{\includegraphics[width=0.7\textwidth]{introduction/images/ai-overview}}
\caption{Dendrogram showing an overview of the field of AI with the position of ML emphasised [source unknown]}
\label{fig:ai-overview}
\end{figure}

\textit{Explainable AI} (xAI) is the sub-field of Artificial Intelligence that ideally should rest at the intersection between Computer Science, Social Sciences and Philosophy and whose aim should be to define the desiderata of artificially intelligent systems and to develop methods to achieve them.
For example, how should a self-driving car behave when confronted with a real-world analogous of the classic Trolley Problem - a situation where each course of action is liable to cause harm?  
On what basis should a person be denied a mortgage, access to university or a job interview?  How can we be sure that there is no bias in the system?  How do we even define if the system is behaving morally?  Would it currently be feasible for a person that feels they have been harmed by such a decision to appeal it?\footnote{For an overview of AI ethics see \citet{Bostrom2011}}
The main idea to achieve these goals is for the systems in question to be somehow made \textit{explainable}. 
The problems start at the very first step, as within the xAI community, as noted by \citet{Doran2018}, there is currently no unanimously agreed upon definition of the definition of \textit{explainability} and consequently of the best way to achieve it in real systems.
The review carried out in Section \ref{sec:explainability} highlights one of the fundamental problems of the field of xAI: that all researchers within the community claim their methods to be \enquote{explainable}, but very few justify this with reasons grounded in the real world (as best noted by the popular paper by \citet{Lipton2016}).

To solve this conundrum various authors have tried to define taxonomies into which classify systems, based on their characteristics and how these relate to its perceived explainability.
One of the most compelling of these is the one proposed by \citet{Doran2018} and shown in Figure \ref{fig:xai-systems-taxonomy}.
The highest level in this classification is taken by \textit{explainable systems} i.e., those that emit explicit, human-understandable reasonings.
This category appears somewhat nebulous in light of the previous paragraph: how should an explainable system be recognised if there is no good definition of explanation?
A less elaborated, but still conceptually sound, taxonomy is the one proposed by \citet{mittelstadt2019explaining} who propose to classify systems based on the source/locus of their explainability.
In this classification models may be \textit{ante-hoc} or \textit{post-hoc} explainable; the former identifies systems whose explainability stems from some internal, inherent characteristic while the latter those whose source of explainability is to be found in some external behaviour, for example the emission of extra symbols.
The two taxonomies are somewhat overlapping even though the latter does not consider \textit{opaque} systems: \textit{interpretable} could be identified with \textit{ante-hoc} while \textit{comprehensible} and \textit{explainable} with \textit{post-hoc}.
A third orthogonal classification by \citet{doshi2017towards} deals with how to \enquote{evaluate an evaluation} of an explanation.
The first two will be presented in more detail in Section \ref{sec:explainability} while the last in Section \ref{sec:evaluation-of-explainability}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.9\textwidth]{introduction/images/xai-systems-taxonomy}}
\caption{The relationship between \textit{opaque}, \textit{interpretable},  \textit{comprehensible} and \textit{interpretable} systems [adapted from \citet{Doran2018}]}
\label{fig:xai-systems-taxonomy}
\end{figure}

What I hope can be gleamed from this brief introduction to the field of Explainable Artificial Intelligence, is that many of the problems it aims to tackle are hard \textit{per-se} and may not have a unique optimal solution.  
This is because these issues are not only engineering problems, but exist at the intersection between man and machine and as such can't be tackled using only the methods of Computer Science (this is what is meant by \citet{doshi2017towards} when they talk about \enquote{incompleteness in the problem formalization}).  
This is one of the main pitfalls that the xAI community finds itself falling into, as most of its researchers come from the hard-sciences domains of computer science, mathematics and statistics. 
There is little hope to define the desiderata without the guidance that can only come from philosophy, because of its millennia-long tradition in dealing with ethical issues.  
There is no way to satisfactorily move towards and evaluate these desiderata without resorting to the well-established literature and methods of the Social Sciences (for a good example of how this may work, see \citet{stumpf2009interacting}), as the human element is inherent in any explanation. 
It would be \enquote{reinventing the wheel} to try and define how a computer should relate itself to its user, when the field of Human Computer Interaction has many of these answers already. 
It should be clear that when the human - and particularly the ethical - domain are part of the equation, it is impossible \textit{by definition} to find an optimal and unique solution, that many xAI researchers seem to still regard as achievable.

Effectively, the biggest gap that can be identified is a dearth of explainability methods that have been validated not only with domain experts, but even with real humans.
Many researchers seem content with claiming \textit{formal explainability} and neglect the human half of the explanation.
An explanation, by its nature, involves and \textit{explainer}, the machine, and an \textit{explainee}, us.
Up till now, it seems that xAI is content with \enquote{explaining the explainer}, not realising that in doing so it offers no explanation at all (see \citet{mittelstadt2019explaining} for a critique of the field of xAI, based on this lack of awareness).

>>>>

\section{Context} \label{sec:intro-context}
While \textit{artificial intelligence} (AI) - as a field - has existed for nearly seventy years \citep{moor2006dartmouth}, the concept of artificial intelligence dates back at least to ancient Greece.  
In early times, artificial intelligence, embodied in the dream of mechanical men, was part of the domain of myth; in the twentieth century, of that of science.  
During this last decade, artificial intelligence is no longer part of a single, specific domain, as it has materialised from Man's imagination, broken out of laboratories and has been given license to act in the world at large.

Artificial intelligence and specifically \textit{machine learning}, the branch of AI that specialises in creating computer systems able to learn their programming from real-world examples instead of having to be explicitly coded by a human, have in the last two decades seen extraordinary success\footnote{See \citet{shalev2014understanding} for an introduction to machine learning and Figure \ref{fig:ai-overview} to get an overview of the field of AI.}.
No sector of our economy has been left untouched by the recent and rapid rise of machine learning, which has been enabled by the rediscovery of neural networks, the availability of big data and increased access to parallel computing power\footnote{A compelling talk, explaining the recent rise of ML, by AI pioneer Geoff Hinton can be found at \url{https://www.youtube.com/watch?time_continue=1&v=izrG86jycck}.}.
Fields as diverse and critical as are government, healthcare, finance and bioinformatics have been revolutionised and the opportunity has been opened for new sectors - such as self-driving vehicles - to appear\footnote{For an overview of the impacts of AI on current society and labour market see \citet{schwab2017fourth}.}.
The increasing reliance of our civilisation on ever more complex machine learning-driven algorithms can only make us more worried about the ethical problems posed by such a series of circumstances.  
Our society has only very recently been confronted with the dilemma of assigning responsibility when a driverless car causes the death of a person\footnote{\url{https://www.nytimes.com/interactive/2018/03/20/us/self-driving-uber-pedestrian-killed.html}}.
This moral issue is only the tip of the iceberg, even when focusing exclusively on the automotive industry.  
As an increasing number of decisions are made in an automated way, with many of them significantly impacting both individuals and society at large, it is natural to stop and wonder which characteristics are desirable in the systems tasked with giving us these verdicts.

 \begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{introduction/images/ai-overview}}
\caption{Dendrogram showing an overview of the field of AI with the position of ML emphasised [\href{https://tu-dresden.de/ing/informatik/ressourcen/dateien/studium/lehrangebot/forschungslinie_einfuehrung_in_die_forschung/2017/2017-04-03-FG-KI.pdf}{Sebastian Rudolph}].}
\label{fig:ai-overview}
\end{figure}

\textit{Explainable AI} (xAI) is the sub-field of artificial intelligence that ideally should be found at the intersection between computer science, social sciences and philosophy and that should aim to define the desiderata of artificially intelligent systems and to develop methods to achieve these.
For example, how should a self-driving car behave when confronted with a real-world predicament analogous to the classic trolley problem - a situation where each course of action is liable to cause harm?  
On what basis should a person be denied a mortgage, access to university or a job interview?  How can we be sure that the bias in the system is reduced to an acceptable level?  How do we even define if the system is behaving morally?  Would it currently be feasible for a person  to appeal an automated decision they feel they had been harmed by?\footnote{For an overview of AI ethics see \citet{Bostrom2011}.}
The main strategy developed in order to achieve these goals is for the systems in question to be somehow made \textit{explainable}. 
The problems in this process start at the very first step, given that within the xAI community, as noted by \citet{Doran2018}, there is currently no unanimously agreed upon definition of \textit{explainability} and consequently of the best way to achieve it in real systems.
The review carried out in Section \ref{sec:explainability} highlights one of the fundamental problems of the field of xAI: that all researchers within the community claim their methods to be \enquote{explainable}, but very few justify this with reasons grounded in the real world (as best summarised in the popular paper \citep{Lipton2016}).

To solve this conundrum, various authors have tried to define taxonomies which classify systems based on their characteristics and how these relate to their perceived explainability.
One of the most compelling of these is that proposed by \citet{Doran2018} and shown in Figure \ref{fig:xai-systems-taxonomy}.
The highest level in this classification is occupied by \textit{explainable systems} i.e., those that emit explicit, human-understandable reasonings.
This category appears somewhat nebulous in light of the previous paragraph: how should an explainable system be recognised if there is no good definition of explanation?
A less elaborate, but still conceptually sound taxonomy is the one proposed by \citet{mittelstadt2019explaining} who propose to classify systems based on the source, or \textit{locus}, of their explainability.
In this classification, models may be \textit{ante-hoc} or \textit{post-hoc} explainable; the former identifies systems whose explainability stems from some internal, inherent characteristic while the latter those whose source of explainability is to be found in some external behaviour, for example in the emission of extra symbols along the output.
The two taxonomies are somewhat overlapping - even though the latter by \citet{mittelstadt2019explaining} does not consider \textit{opaque} systems; the two classifications could be put into relationship by identifying \textit{interpretable} with \textit{ante-hoc} and \textit{comprehensible} together with \textit{explainable} with \textit{post-hoc}.
A third orthogonal classification forwarded by \citet{doshi2017towards} addresses how to evaluate the quality of an explanation.
The first two taxonomies will be presented in more detail in Section \ref{sec:explainability} while the last in Section \ref{sec:evaluation-of-explainability}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.9\textwidth]{introduction/images/xai-systems-taxonomy}}
\caption{The relationship between \textit{opaque}, \textit{interpretable},  \textit{comprehensible} and \textit{interpretable} systems [adapted from \citep{Doran2018}].}
\label{fig:xai-systems-taxonomy}
\end{figure}

The purpose of this brief introduction to explainable artificial intelligence, is to help in understanding that many of the problems that the field aims to tackle are hard \textit{per-se} and may not have a unique optimal solution.  
This is because such issues are not only engineering problems, but exist at the intersection between man and machine and as such cannot be solved using only engineering methods\footnote{This is what is meant by \citet{doshi2017towards} when they talk about \enquote{incompleteness in the problem formalization}.}.  
This lack of awareness for the centrality of the human element of an explanation is one of the main pitfalls of the xAI community, since most of its researchers come from the \textit{hard-science domains} of computer science, mathematics and statistics. 
There is little hope of defining the desiderata for intelligent systems without the guidance that can only come from philosophy, because of its millennia-long tradition in dealing with ethical issues.  
There is also no way to satisfactorily move towards and evaluate these desiderata without resorting to the well-established literature and methods of the social sciences\footnote{For a good example of how this may work, see \citet{stumpf2009interacting}.}, as the human element is inherent in any explanation. 
An attempt to define how a computer should relate itself to its user could be assimilated with the notion of \enquote{reinventing the wheel}, as the field of human computer interaction has many such techniques already. 
Although an endeavour still pursued by many xAI researchers, it should be clear that when the human - and particularly the ethical - domain are part of the equation, it is impossible \textit{by definition} to find an optimal and unique solution\footnote{See the concept of \enquote{wicked problem} in \citep{Rittel1973}.}.

Effectively, the biggest gap that can be identified is a dearth of explainability methods that have been validated not only by domain experts, but even by real humans.
Many researchers seem content with claiming \textit{formal explainability} and neglect the human element of the explanation.
An explanation, by its vert nature, involves an \textit{explainer}, the machine, and an \textit{explainee}, us humans.
Up till now, it seems that xAI is content with only explaining the machine but, in doing so, overlooks the fact that it may be offering the users no explanation at all\footnote{See \citet{mittelstadt2019explaining} for a critique of the field of xAI, based on this lack of awareness.}.
